---
title: " Teaching Data Science"
subtitle: "<br/>Chapter 16"
author: "Joshua Rosenberg"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
  xaringan::moon_reader:
    # css: ["default", "metropolis", "metropolis-fonts"]
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
      ratio: '16:9'
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono")
)
```

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

class: middle

# Learning Objectives

1. Pedagogical features embedded within the book
1. Teaching data science
1. Overcoming barriers to doing data science

---

class: middle

# But, before that

## In response to the question, 'How should I get started doing data science'?

### What would you recommend they do first, second, and third (subsequently) to get started?

---

# Pedagogical principles for the book

- Problem-based learning
  - Supporting learning around real-world challenges
  - The _walkthroughs_ were designed with actual problems edu. data scientists face in mind
- Differentiation
  - Chapters 5 and 6 for those completely new to R
  - Appendices as 'reaches'
- Working in the open
  - [R View blog post](https://rviews.rstudio.com/2020/07/01/open-source-authorship-of-data-science-in-education-using-r/)
- Building mental models
  - Foundational Skills Framework: Projects, functions, packages, data
- Universal design
  - Inclusive and accessible to all individuals
  - Did not adequately address (based upon our aim)
  
---

class: middle

# What was helpful to you? 
# What additional (helpful features could be added or created)?

---

# Strategies for teaching data science

- Provide a home base
  - Reduce some of the work that learners have to do that is unrelated to reasoning about and using code
- Write code early and often
  - Most everyone can learn to code; begin with reasoning about the output of code and writing code early
- Don't touch that keyboard!
  - Supporting learners to fix typos and become familiar with syntax is well-worth the effort
- Anticipate issues (and sacrifice accuracy for clarity early on)
  - The curse of knowledge is hard to overcome; be _generally_ accurate but emphasize being clear and pragmatic early on
- Start with "early wins"
  - It is important to consider the end point; what will learners _want_ to do or make?
- Consider representation and inclusion
  - Doing data science and working with data in education make these _even more_ important than in other domains in that we can make impressions upon learners about who does data science and what data we create 
- Draw on other resources
  - There are great resources in computing education and K-12 education to draw on

---

class: middle

# What are other helpful teaching strategies?
# How do these strategies look different with learners of different ages and in different contexts?

---

# What's next?

- There is a lot of growth and interest in data science education
- But, learning to do data science is still _really hard_!
- **What are big gaps in resources?**

--- 

# Thank you!

Joshua Rosenberg  
jmrosenberg@utk.edu  
@jrosenberg6432