---
title: " Teaching Data Science"
subtitle: "<br/>Chapter 16"
author: "Joshua Rosenberg"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
  xaringan::moon_reader:
    # css: ["default", "metropolis", "metropolis-fonts"]
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
      ratio: '16:9'
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono")
)
```

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

class: middle

# Learning Objectives

1. Where would you begin?
1. Discuss strategies for teaching data science
1. Discuss some general strategies related to teaching and learning
1. Discuss the pedagogical principles for the book
1. Discuss teaching science across ages and contexts
1. Discuss big gaps in resources

---

# Where would you begin when teaching someone else?

---

# Strategies for teaching data science

- Provide a home base
- Write code early and often
- Don't touch that keyboard!
- Anticipate issues (and sacrifice accuracy for clarity early on)
- Start with "early wins"
- Consider Representation and Inclusion in the Data and Examples You Use
- Draw on other resources

---

# General strategies related to teaching and learning

- Recognize that learning is complex
- Learners learn many things at once
- Meta-cognition is important
- Learning strategies can make a difference

---

# Pedagogical principles for the book

- Problem-based learning
- Differentiation
- Working in the open
- Building mental models
- Universal design

---


# Teaching science across ages and contexts

- 
- 
- 
- 

--- 

# Where are big gaps in resources?

- 
- 
- 
- 

--- 

# Thank you!

Joshua Rosenberg  
jmrosenberg@utk.edu  
@jrosenberg6432
