---
title: "DSiEUR Bookclub - Ch. 5 & 6"
author: "Mark LaVenia"
date: "3/3/2021"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Useful Links

Data Science in Education Using R
Ryan A. Estrellado, Emily A. Bovee, Jesse Mostipak, Joshua M. Rosenberg, and Isabella C. Velásquez
<https://datascienceineducation.com/>  
  
    
Bookclub GitHub Repo
<https://github.com/r4ds/bookclub-dsieur>

## Chapters and Topics

- Ch 5 Getting Started with R and RStudio
- Ch 6 Foundational Skills

## 5.6.2 Writing Code in an R Script

print("We're going to use R as a calculator.")  
print("First up, addition!")  
12 + 8  
632 + 41   
print("Next, subtraction!")   
48 - 6  
0.65 - 1.42

## R Script returned results

```{r}

print("We're going to use R as a calculator.")
print("First up, addition!")
12 + 8
632 + 41
print("Next, subtraction!")
48 - 6
0.65 - 1.42

```

## 5.7 Installing the {dataedu} Package


- The recommended method for downloading dataedu kept giving me an error message
  - install.packages("devtools", repos = "http://cran.us.r-project.org")
  - devtools::install_github("data-edu/dataedu")
<br><br>
- So I went to the repo <https://github.com/data-edu/dataedu>, where it recommended using the "remotes" package, which worked for me.
  - install.packages("remotes")
  - remotes::install_github("data-edu/dataedu")

## 6.5 Foundational Skills Framework

- Projects  
- Functions  
- Packages  
- Data  

## Writing a function:
'# we've named the function "addition" and asked for two arguments, "number_1" and "number_2"  
addition <- function(number_1, number_2) {  
    number_1 + number_2  
}  

addition(number_1 = 3, number_2 = 1)  

addition(0.921, 12.01)  

addition(62, 34)

## Function output

```{r}
#' writing our function
#' we've named the function "addition"
#' and asked for two arguments, "number_1" and "number_2"
addition <- function(number_1, number_2) {
    number_1 + number_2
}

#' using our function
#' below are 3 separate examples of utilizing our new function called "addition"
#' note that we provide each argument separated by commas
addition(number_1 = 3, number_2 = 1)

addition(0.921, 12.01)

addition(62, 34)

```

## Interpretting code

library(tidyverse) 
library(janitor)  

roster <- roster_raw %>%   
    clean_names() %>%   
    remove_empty(c("rows", "cols")) %>%  
    mutate(hire_date = excel_numeric_to_date(hire_date),  
           cert = coalesce(certification, certification_1)) %>%   
    select(-certification, -certification_1)  

## apply filter function

`# using the filter() function from the stats package  
x <- 1:100  

stats::filter(x, rep(1, 3))  

`# using the filter() function from the dplyr package  
starwars %>%   
    dplyr::filter(mass > 85)
    
## Loading data

dataedu::ma_data_init

dataedu::ma_data_init -> ma_data

ma_data_init <- dataedu::ma_data_init

## Exploring and manipulating your data

names(ma_data_init)  

glimpse(ma_data_init)  

summary(ma_data_init)  

glimpse(ma_data_init$Town)  

summary(ma_data_init$Town)    

glimpse(ma_data_init$`AP_Test Takers`)  

summary(ma_data_init$`AP_Test Takers`)  

## 6.12.4 Exploring Our Data with the Pipe Operator

ma_data_init %>%  
    group_by(`District Name`) %>%  
    count()  
    
ma_data_init %>% 
    group_by(`District Name`) %>%  
    count() %>%  
    filter(n > 10) 

## Exploring Our Data with the Pipe Operator Output (cont.)

ma_data_init %>%  
    group_by(`District Name`) %>%  
    count() %>%  
    filter(n > 10) %>%  
    arrange(desc(n)) 

```{r}





```

